---
title: "Isolation-by-distance: Significance of genetic distance comparisons"
author: "Jayna Bergman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Version of R used: `r getRversion()`
<br> 

## Project

Western toad DUs Genomics. 
<br> 
<br>


## Overview

### **Goal(s):** 
The goal of this script is to test for significance between the genetic differences when comparing calling vs non-calling to within each group
<br>

### **Details about files and pipeline:** 
The input dataframe was generated in the previous IBD script (where Nei's genetic distance and geographic distances (m) was calculated between each pair of individuals). The dataframe does not include individuals from the "southern" population (i.e. no WLNP or Corbin individuals)

<br>
<br>

## Setting up the R environment

### Working directory: 

```{r, message =FALSE}
setwd("C:/Users/jayna/OneDrive/Desktop/School/Lee_Yaw_Lab/Masters_Thesis/Data_chapter/RMD")
```

### Libraries: 

```{r, message=FALSE}
# load libraries
library(dplyr)
library(knitr)
library(ggplot2)
library(tidyverse)
library(tidyr)
library(vcfR)
library(adegenet)
library(poppr)
library(hierfstat)
library(ape)
library(ade4)
library(dartR)
library(geosphere)
library(patchwork)
```

## **Step 1 of analysis:** 

### Load and explore the data

Load the matrix that was made in the script before:

```{r}
data <- read.csv("C:/Users/jayna/OneDrive/Desktop/School/Lee_Yaw_Lab/Masters_Thesis/Data_chapter/RMD/IBD/Proportion_shared_alleles/All_ind_IBD_matrix_wo_WLNP_PropSA.csv")
```

Look at the summary of the data:

```{r}
summary(data)
```

Subset and filter the data to get the columns for the dataframe:

```{r}
## Select columns to keep
data_subset <- data %>%
  select(Gen_dist, Geog_dist, Type)
  
## Subset for 150km between points - standardizing geographic distances  
data_150km <- data_subset %>%
  filter(Geog_dist <= 150)
```

## **Step 2 of analysis:** All points

Run an ANOVA on the genetic distance and Type (for all points):

```{r}
## Run anova test
Test <- aov(Gen_dist ~ Type, data = data_subset)

Test

data_results <- summary(Test)
data_results

## Save F and P values
data_f_value <- data_results[[1]]$`F value`[1]
data_p_value <- data_results[[1]]$`Pr(>F)`[1]
```

Determine the mean genetic distance of each comparison type:

```{r}
## group by type and calculate the mean of each type
real_results <- data_subset %>%
    group_by(Type) %>%
    summarise(mean_gendist = mean(Gen_dist))%>% 
    pivot_wider(names_from = Type, values_from = mean_gendist)

real_results <- real_results %>%
  mutate(diff_cvc_cvnc = `Calling vs Calling` - `Calling vs Non-calling`,
         diff_nvn_cvnc = `Non-calling vs Non-calling` - `Calling vs Non-calling`)

print(real_results)

## Add additional columns from the ANOVA to the dataframe - name it "real data"
real_results$F_value <- data_f_value
real_results$P_value <- data_p_value
real_results$Iteration <- "real data"
```

Make plot to show to real data points and the mean:

```{r}
## Saved with the mean point as one version and the boxplot as a second version

A_plot <- ggplot(data_subset, aes(x = Type, y = 1-Gen_dist, col = Type)) +
  geom_point(shape = 16, position = position_jitter(w=0.1, h=0)) + 
  geom_boxplot(outlier.shape = NA, color = "black", fill = NA) +
  #stat_summary(fun = mean, geom = "point", size=2.5, color = "black") +
  scale_color_manual(values = c("Calling vs Non-calling"= "#CAB2D6", "Calling vs Calling" = "#FDBF6F", "Non-calling vs Non-calling"= "#A6CEE3")) +
  labs(y = "1 - Proportion shared alleles", x = "Comparison Type") +
  theme(legend.position = "none", 
        axis.text.x = element_text(size=10),
        axis.text.y = element_text(size= 10),
        panel.background=element_rect(fill="white", color=NA),
        panel.grid.major = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.line = element_line(color="black"),
        axis.ticks.x=element_blank(),
        axis.title.x = element_text(margin=margin(t=17)),
        axis.title.y = element_text (margin=margin(r=7), size=11)) 

A_plot

ggsave(plot=A_plot, "C:/Users/jayna/OneDrive/Desktop/School/Lee_Yaw_Lab/Masters_Thesis/Data_chapter/RMD/IBD/Proportion_shared_alleles/IBD_ANOVA_allpts_boxplot.jpeg", dpi=800, width=8, height=8)
```


Randomizing one of the columns then taking the average genetic distance of each comparison type:

```{r}
# Create empty list 
results_list <- list()

# Number of iterations
n_iterations <- 1000

for (i in 1:n_iterations) {
  ## Randomize the type column
  data_subset$Randomized_type <- sample(data_subset$Type)
  
  ## Run ANOVA
  anova_result <- aov(Gen_dist ~ Randomized_type, data = data_subset)
  anova_summary <- summary(anova_result)
  
  F_value <- anova_summary[[1]]$`F value`[1]
  P_value <- anova_summary[[1]]$`Pr(>F)`[1]
  
  ## group by type and calculate the mean of each type
  mean_results <- data_subset %>%
    group_by(Randomized_type) %>%
    summarise(mean_gendist = mean(Gen_dist)) %>%
    pivot_wider(names_from = Randomized_type, values_from = mean_gendist)
    
  ## Calculate differences between the means
  mean_results <- mean_results %>%
  mutate(diff_cvc_cvnc = `Calling vs Calling` - `Calling vs Non-calling`,
         diff_nvn_cvnc = `Non-calling vs Non-calling` - `Calling vs Non-calling`)
  
  ## Add additional columns to the new dataframe
  mean_results$F_value <- F_value
  mean_results$P_value <- P_value
  mean_results$Iteration <- i
 
  
  ## Store results
  results_list[[i]] <- mean_results
  
}

head(results_list[100])

## Combine the different iterations
final_results <- do.call(rbind, results_list)

## Add on the real results
final_results <- rbind(final_results, real_results)

write.csv(final_results, "C:/Users/jayna/OneDrive/Desktop/School/Lee_Yaw_Lab/Masters_Thesis/Data_chapter/RMD/IBD/ANOVA_mean_results_differences.csv")
```

Make a histogram distribution of the means and where the real data mean is:

```{r}
##### Plot for calling vs calling and calling vs non-calling 
## Sub set data
calling_calling <- final_results %>%
  select(diff_cvc_cvnc, Iteration)

## Setting the value to add the line of the "real data" point
Real_data_C <- "real data"
iteration_diffs_C <- calling_calling$diff_cvc_cvnc[calling_calling$Iteration == Real_data_C]


plot_hist1 <- ggplot(calling_calling, aes(x=diff_cvc_cvnc)) +
  geom_histogram(binwidth = 0.001, fill = "#CAB2D6", color = "black", alpha = 0.7) +
  geom_segment(aes(x = iteration_diffs_C, xend = iteration_diffs_C, y = 0, yend = 75), color = "red", linetype = "dashed", size = 0.8) +
  scale_y_continuous(expand = c(0,0), limits=c(0,75), breaks=seq(0, 75, by= 25)) +
  labs(title = " ", x = "Difference between the mean genetic distances (c vs c and c vs nc)", y = "Frequency") +
  theme(legend.position="none", 
        axis.text.x = element_text(size=12),
        axis.text.y = element_text(size= 12),
        panel.background=element_rect(fill="white", color=NA),
        panel.grid.major = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.line = element_line(color="black"),
        axis.ticks.x=element_blank(),
        axis.title.x = element_text(margin=margin(t=17), size=14),
        axis.title.y = element_text (margin=margin(r=7), size=14))


plot_hist1

ggsave(plot=plot_hist1, "C:/Users/jayna/OneDrive/Desktop/School/Lee_Yaw_Lab/Masters_Thesis/Data_chapter/RMD/IBD/calling_noncalling_mean_histogram_differences.jpeg", dpi=800, width=10, height=8)



##### Plot for noncalling vs noncalling and calling vs noncalling

## Subset data
noncalling_noncalling <- final_results %>%
  select(diff_nvn_cvnc, Iteration)

## Setting the value to add the line of the "real data" point
Real_data_NC <- "real data"
iteration_diffs_NC <- noncalling_noncalling$diff_nvn_cvnc[noncalling_noncalling$Iteration == Real_data_NC]


plot_hist2 <- ggplot(noncalling_noncalling, aes(x=diff_nvn_cvnc)) +
  geom_histogram(binwidth = 0.001, fill = "#CAB2D6", color = "black", alpha = 0.7) +
  geom_segment(aes(x = iteration_diffs_NC, xend = iteration_diffs_NC, y = 0, yend = 75), color = "red", linetype = "dashed", size = 0.8) +
  scale_y_continuous(expand = c(0,0), limits=c(0,75), breaks=seq(0, 75, by= 25)) +
  labs(title = " ", x = "Difference between the mean genetic distances (nc vs nc and c vs nc)", y = "Frequency") +
  theme(legend.position="none", 
        axis.text.x = element_text(size=12),
        axis.text.y = element_text(size= 12),
        panel.background=element_rect(fill="white", color=NA),
        panel.grid.major = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.line = element_line(color="black"),
        axis.ticks.x=element_blank(),
        axis.title.x = element_text(margin=margin(t=17), size=14),
        axis.title.y = element_text (margin=margin(r=7), size=14))


plot_hist2

ggsave(plot=plot_hist2, "C:/Users/jayna/OneDrive/Desktop/School/Lee_Yaw_Lab/Masters_Thesis/Data_chapter/RMD/IBD/calling_calling_mean_histogram_differences.jpeg", dpi=800, width=10, height=8)
```

Make patchwork figure:
```{r}
Figure_allpts <- (plot_hist1) /
  (plot_hist2) +
  plot_annotation(tag_levels = 'A',
                  tag_suffix = ")")

Figure_allpts

ggsave(plot=Figure_allpts, "C:/Users/jayna/OneDrive/Desktop/School/Lee_Yaw_Lab/Masters_Thesis/Data_chapter/RMD/IBD/All_points_mean_histogram_differences.jpeg", dpi=800, width=10, height=12)
```

### Conclusions 

The differences between the means are statistically larger than expected by the null distribution.

<br>
<br> 

### Specific notes for the manuscript/thesis 

The first anova plot will go into the main manuscript and the histograms will go into the sup mat. 

<br>
<br> 

### References 
